Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-08-07 docs/papers/shallow_performance/shallow_perf.tex
Add NERC grant no. to acks

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Towards Compiler-Agnostic Performance in Finite-Difference Codes}
%
% author names and IEEE memberships
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{ A.~R.~Porter, M.~Ashworth, R.~W.~Ford, M.~Modani, G.~D.~Riley% <-this % stops a space
\thanks{A.~R.~Porter, M.~Ashworth, R.~W.~Ford and M.~Modani work in the Science and Technology Facilities Council's Daresbury Laboratory, UK.}% <-this % stops a space
\thanks{G.~D.~Riley works in the School of Computer Science at the University of Manchester, Manchester, UK.}
}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names.

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~11, No.~4, December~2012}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2012 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique called \psykal to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion etc) are then
applied to the \psykal version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem sizes
apart from GNU on Haswell, one problem size for Intel on Haswell and
IBM on Power8.
%
As the GNU compiler performs well on Ivybridge the reduction in
performance for the \psykal version was not analysed further. As the
Intel performance on Haswell was better for the \psykal version than
the original version, it was not analysed further.
%
In the case of the IBM compiler, the \psykal performance is
significantly better than the original hand optimised code on all
problem sizes. Analysing the performance of the IBM compiler shows
that a different structure for the original optimised code produces
much improved performance. Replicating this different structure using
appropriate transformations in the \psykal version provides the
same performance improvements.
%
Therefore, the \psykal approach can be used with negligable
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the PSyKAl version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Performance, Code-generation, Finite-difference
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HE" in caps to complete the first word.
\IEEEPARstart{T}{he} challenge presented to the developers of
scientific software by the drive towards Exascale computing is
considerable. With power consumption becoming the overriding design
constraint, CPU clock speeds are falling and the complex,
multi-purpose compute core is being replaced by multiple, simpler
cores. This philosophy can be seen at work in the rise of so-called
accelerator based machines in the Top 500 List~\cite{top500} of
supercomputers: five of the top-ten machines in the November 2014 list
make use of Intel Xeon Phi's or NVIDIA GPUs. Four of the remaining
five machines in the top ten are IBM BlueGene/Qs, the CPU of which has
hardware support for running 64 threads.

@@

% needed in second column of first page if using \IEEEpubid
%\IEEEpubidadjcol

@@

%\begin{figure}
%\centering
%\caption{The \psykal Approach.}
%\end{figure}

@@

% \bf{Perhaps we don't want to give too much away here so I've removed the optimisation space search plans even though it is mentioned in the text}
% Finally we plan to use this compiler to allow us to search the optimisation sp%ace of transformations to automatically optimise a code for a particular compil%er/architecture combination or to determine whether we can find improved perfor%mance over existing hand optimised solutions and to automatically

@@

This work made use of the ARCHER UK National Supercomputing Service
(\url{http://www.archer.ac.uk}). This work was funded by NERC project {\bf xxx}

% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,shallow_perf}
%

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

\begin{IEEEbiographynophoto}{A.~R.~Porter}
Biography text here.
\end{IEEEbiographynophoto}

\begin{IEEEbiographynophoto}{M.~Ashworth}
Biography text here.
\end{IEEEbiographynophoto}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{R.~W~Ford}
Biography text here.
\end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

\begin{IEEEbiographynophoto}{M.~Modani}
Biography text here.
\end{IEEEbiographynophoto}

\begin{IEEEbiographynophoto}{G.~D.~Riley}
Biography text here.
\end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\pagestyle{headings}
\def\thepage{}

\begin{frontmatter}              % The preamble begins here.
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Towards Compiler-Agnostic Performance in Finite-Difference Codes}

\author[A]{\fnms{A. R.} \snm{Porter}%
\thanks{Corresponding Author: Andrew Porter, STFC Daresbury Laboratory, Warrington, WA4 4AD, UK; E-mail:
andrew.porter@stfc.ac.uk.}},
\author[A]{\fnms{R. W.} \snm{Ford}},
\author[A]{\fnms{M.} \snm{Ashworth}},
\author[B]{\fnms{G. D.} \snm{Riley}}
and
\author[C]{\fnms{M.} \snm{Modani}}
\runningauthor{A.~R.~Porter et al.}
\address[A]{STFC Daresbury Laboratory, Warrington, WA4 4AD, UK}
\address[B]{The University of Manchester, UK}
\address[C]{STFC Hartree Centre}

\begin{abstract}
In this paper we evaluate the performance implications of applying a
technique called \psykal to finite difference Ocean models. 
%
In \psykal, the code related to the underlying science is formally
separated from code related to parallelisation and single core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
(thereby keeping a single code base) and for optimisation specialists
to be able to tailor the code for a particular machine independently
of the science code.
%
A finite difference shallow water benchmark optimised for cache based
architectures is taken as the starting point. A vanilla \psykal version
is written and the performance of the two compared. The optimisations
that were applied to the original benchmark (loop fusion etc) are then
applied to the \psykal version as a set of code modifications to the
optimisation layer. Performance results are presented for the Cray,
Intel and GNU compilers on Intel Ivybridge and Haswell processors and
for the IBM compiler on Power8.
%
Results show that the combined set of code modifications obtain
performance that is within a few percent of the original code for all
compiler and architecture combinations on all tested problem sizes
apart from GNU on Haswell, one problem size for Intel on Haswell and
IBM on Power8.
%
As the GNU compiler performs well on Ivybridge the reduction in
performance for the \psykal version was not analysed further. As the
Intel performance on Haswell was better for the \psykal version than
the original version, it was not analysed further.
%
In the case of the IBM compiler, the \psykal performance is
significantly better than the original hand optimised code on all
problem sizes. Analysing the performance of the IBM compiler shows
that a different structure for the original optimised code produces
much improved performance. Replicating this different structure using
appropriate transformations in the \psykal version provides the
same performance improvements.
%
Therefore, the \psykal approach can be used with negligable
performance loss and sometimes small performance gains compared to the
original optimised code, there is no single best hand optimised
implementation of the code for all the compilers tested, and the
flexibility in applying code modifications to the optimisation layer
in the PSyKAl version allows for a significant performance improvement
on the IBM Power8 compared with the original code.

@@

\begin{keyword}
Performance, Code-generation, Finite-difference
\end{keyword}

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction is not to be numbered so use \section*{}
\section*{Introduction}
The challenge presented to the developers of
scientific software by the drive towards Exascale computing is
considerable. With power consumption becoming the overriding design
constraint, CPU clock speeds are falling and the complex,
multi-purpose compute core is being replaced by multiple, simpler
cores. This philosophy can be seen at work in the rise of so-called
accelerator based machines in the Top 500 List~\cite{top500} of
supercomputers: five of the top-ten machines in the November 2014 list
make use of Intel Xeon Phi's or NVIDIA GPUs. Four of the remaining
five machines in the top ten are IBM BlueGene/Qs, the CPU of which has
hardware support for running 64 threads.

@@

@@

@@

@@

This work made use of the ARCHER UK National Supercomputing Service
(\url{http://www.archer.ac.uk}). We acknowledge use of the Hartree
Centre IBM Power 8 system in this work. The STFC Hartree Centre is a
research collaboratory in association with IBM providing High
Performance Computing platforms funded by the UK's investment in
e-Infrastructure. The Centre aims to develop and demonstrate next
generation software, optimised to take advantage of the move towards
exa-scale computing. This work was funded by the NERC GOcean project,
grant number NE/L012111/1.

% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
\bibliographystyle{unsrt}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{shallow_perf}
%
\end{document}
